\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
created Makefile to meet RJournal requirments and make a zip file

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given their covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
  \newcommand{\proglang}{\textsf}
output: rticles::rjournal_article 
bibliography: casebase-manuscript.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

```{r, eval = FALSE, echo=FALSE}
print.summary.glm <- function(x,digits=max(3,getOption("digits")-3),symbolic.cor=x$symbolic.cor,
                              signif.stars = getOption("show.signif.stars"),show.call=FALSE,show.resids=FALSE,...) {
  if (show.call) cat("Call:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"), "\n", sep = "")
  if (show.resids) {
    cat("Deviance Residuals: \n")
    if (x$df.residual > 5) {
      x$deviance.resid <- stats::quantile(x$deviance.resid, na.rm = TRUE)
      names(x$deviance.resid) <- c("Min", "1Q", "Median", "3Q", "Max")
    }
    xx <- zapsmall(x$deviance.resid, digits + 1)
    print.default(xx, digits = digits, na.print = "", print.gap = 2)
    cat("\n")
  }
  if (length(x$aliased) == 0L) cat("No Coefficients\n")
  else {
    df <- if ("df" %in% names(x)) x[["df"]]
    else NULL
    if (!is.null(df) && (nsingular <- df[3L] - df[1L])) 
      cat("Coefficients: (", nsingular, " not defined because of singularities)\n", sep = "")
    else cat("\nCoefficients:\n")
    coefs <- x$coefficients
    if (!is.null(aliased <- x$aliased) && any(aliased)) {
      cn <- names(aliased)
      coefs <- matrix(NA, length(aliased), 4L, dimnames = list(cn,colnames(coefs)))
      coefs[!aliased, ] <- x$coefficients
    }
    stats::printCoefmat(coefs, digits = digits, signif.stars = signif.stars,na.print = "NA", ...)
  }
  cat("\n(Dispersion parameter for ", x$family$family, " family taken to be ", 
      format(x$dispersion), ")\n\n", apply(cbind(paste(format(c("Null","Residual"), 
                                                              justify = "right"), "deviance:"),
                                                 format(unlist(x[c("null.deviance","deviance")]), digits = max(5, digits + 1)), " on", 
                                                 format(unlist(x[c("df.null", "df.residual")])), " degrees of freedom\n"),1L,
                                           paste, collapse = " "), sep = "")
  if (nzchar(mess <- stats::naprint(x$na.action))) cat("  (", mess, ")\n",sep="")
  cat("AIC: ",format(x$aic,digits=max(4,digits+1)),"\n\n",
      "Number of Fisher Scoring iterations: ",x$iter,"\n",sep="")
  correl <- x$correlation
  if (!is.null(correl)) {
    p <- NCOL(correl)
    if (p > 1) {
      cat("\nCorrelation of Coefficients:\n")
      if (is.logical(symbolic.cor) && symbolic.cor) print(stats::symnum(correl,abbr.colnames=NULL))
      else {
        correl <- format(round(correl, 2),nsmall=2,digits=digits)
        correl[!lower.tri(correl)] <- ""
        print(correl[-1,-p,drop=FALSE],quote=FALSE)
      }
    }
  }
  cat("\n")
  invisible(x)
}

print(summary(fits[["splines"]]), signif.stars = F)
```

@@

# dev.off()
layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xmax = max(new_time),
     # xlim = range(new_time), 
     conf.int = FALSE, col = myCols, 
     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
     #                exp(coef(cox_model)), 
     #                exp(confint(cox_model))[1], 
     #                exp(confint(cox_model))[2]), 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     # xlim = c(0,15), 
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
      xaxt = 'n', yaxt = 'n',
     # xlim = c(0,15), 
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     # xlim = c(0,15), 
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given their covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | * SRB and MT contributed equally to this work.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through three different case studies.
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
  \newcommand{\proglang}{\textsf}
output: rticles::rjournal_article 
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

@@

layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
      xaxt = 'n', yaxt = 'n',
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
created Makefile to meet RJournal requirments and make a zip file

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given their
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis of
time-to-event data. We illustrate all these features through three
different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

For the third case study, we show how \CRANpkg{casebase} can also be
used for variable selection through regularized estimation of the hazard
function as given by \eqref{eq:penest}. To illustrate this
functionality, we use the dataset from the Study to Understand Prognoses
Preferences Outcomes and Risks of Treatment (SUPPORT)
\citep{knaus1995support}. The SUPPORT dataset tracks death in five
American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \CRANpkg{casebase} package. In the
comparisons below, all covariates except \code{sps} and \code{aps} were
used. These two variables correspond to scores for predicting the
outcome that were developed as part of the original study. For more
information about this dataset, the reader is encouraged to look at the
documentation in our package.

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g.~\CRANpkg{SmoothHazard},
\CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account
for it. Again, we hope to include any resulting methodology as part of
this package.

@@

\bibliography{casebase-manuscript.bib}

>>>>

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given their
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
* SRB and MT contributed equally to this work.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis of
time-to-event data. We illustrate all these features through three
different case studies.
}

@@

For the third case study, we show how \CRANpkg{casebase} can also be
used for variable selection through regularized estimation of the hazard
function as given by \eqref{eq:penest}. To illustrate this
functionality, we use the dataset from the Study to Understand Prognoses
Preferences Outcomes and Risks of Treatment (SUPPORT)
\citep{knaus1995support}. The SUPPORT dataset tracks death in five
American hospitals within individuals who are considered seriously ill.
The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \CRANpkg{casebase} package. In the
comparisons below, all covariates except \code{sps} and \code{aps} were
used. These two variables correspond to scores for predicting the
outcome that were developed as part of the original study. For more
information about this dataset, the reader is encouraged to look at the
documentation in our package.

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g. \CRANpkg{SmoothHazard},
\CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account
for it. Again, we hope to include any resulting methodology as part of
this package.

@@

\bibliography{bhatnagar-turgeon-islam-hanley-saarela.bib}
